---
title: "P8105 - HW2"
output: github_document
date: "Due: 2024-10-02"
---

Author: Ravi Brenner (irb2118)

# Introduction
This Rmarkdown document is for P8105 - Data Science 1, homework 2.

# Methods
The data for this homework comes from the course website, but all files can be found in the `data/` folder. Additionally, we will be using packages from the `tidyverse`

```{r, message = FALSE}
library(tidyverse)
```


# Problems

## Problem 1

This problem is about the NYC subway system.

First, we must read in the data, clean up the column names, and select and transform the columns that we need:

```{r}
subway_df <-
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada
  ) |>
  mutate(entry = case_match(entry,
                            "YES" ~ TRUE,
                            "NO" ~ FALSE))
```


Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?

Answer the following questions using these data:

How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here.
How many stations are ADA compliant?
What proportion of station entrances / exits without vending allow entrance?
Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?



## Problem 2

## Problem 3

# Conclusion